---
title: "Lab 11b: Regression data exploration - scatterplots"
author: "Nathan Brouwer  brouwern@gmail.com  @lobrowR"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Merging w/ dplyr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE,fig.height = 3, fig.width = 5)
```

## Lab 11 Part I continued: Data exploration with scatterplots


## References
Meredith et al 1991 Repeated measures experiments in forestry: focus on analysis of response curves.  Can. J. For. Res.




## Load data

If you haven't already, load data_long.csv and turn conc.AL and week to factor variables.  This won't be necessary if you are continuing directly from the previous handout.
```{r}
data.long <- read.csv(file = "data_long.csv")
data.long$conc.AL.FAC <- factor(data.long$conc.AL)
data.long$week.FAC <- factor(data.long$week)
```



## Plotting regression data: scatterplots


### Basic scatter plot for regression

Plot regression-style data with the ggpubr function ggscatter()


```{r}
library(ggpubr)

ggscatter(data = data.long,
          y = "height",
          x = "week")
```



### Change x and y labels

Plot regression-style data with the ggpubr function ggscatter()


```{r}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling")
```



### Change color based on treatment

We can use color = "conc.AL" to color code the data points

```{r}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          color = "conc.AL")
```


"conc.AL" is numeric data and when ggplot and/or ggpubr use a numeric variable to set colors they change the shade gradually from dark to light blue.  Since we have only a 4 different amounts of AL used in the study it makes more sense to use AL as a factor.  Change  color = "conc.AL" to  color = "conc.AL.FAC" to produce this graph

```{r echo=FALSE}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          color = "conc.AL.FAC")
```

### Change shape based on treatment

We can chane the shape also to make to even clearr what the treatmetns are.  Setting shape = "conc.AL.FAC"  will produce the following graph

```{r echo=FALSE}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          color = "conc.AL.FAC",
          shape = "conc.AL.FAC" ) 
```



## Facet data

There is a lot of overlap in the points.  We can clean things up by faceting the data so that the treatments are in different panels.

```{r}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          color = "conc.AL.FAC",
          shape = "conc.AL.FAC",
          facet.by = "conc.AL.FAC")
```


## Add smoothers & regression lines

### Add smoothers lines

Smoothers show you the general shape of the data.  A common type of smoother is a "loess smoother."  We can add one by adding add = "loess"


```{r, warning=FALSE}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          add = "loess") 

```

What happens when we add color?
```{r, warning=FALSE}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          add = "loess",
          color = "conc.AL.FAC") 

```

Most of these lines are fairly straight, meaning that standard regression will probably work pretty well.  When smoother lines curve more complicate models might be appropriate.



### Add regression lines

A least square regression line can be added to the plot using add = "reg.line" for "Add regression line."  This line is fit behind the scenes using the lm() function.

```{r, warning=FALSE}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          add = "reg.line") 

```

What happens when we add color using color = "conc.AL.FAC"?
```{r}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          add = "reg.line",
          color = "conc.AL.FAC") 

```

And then we facet?
```{r, warning=FALSE}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          add = "reg.line",
          color = "conc.AL.FAC",
          facet.by = "conc.AL.FAC") 

```


### Add confidence intervals

Data in a scatterplot are used to estimate an intercept and a slope for the best fit line running through the scatterplot.  The slope and intercept are estimated with uncertainty; therefore the slope and intercept both have standard errors.  This uncertainty can be transformed into an error band or 95% confidence band that represents uncertainty in the true location of the line.

We can easily add this "conf.int = TRUE". Each line has its own 95% confidence interval.  
```{r}
ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          add = "reg.line",
          conf.int = TRUE,
          color = "conc.AL.FAC",
          facet.by = "conc.AL.FAC") 
```


## Advanced / Optional: Combine scatter plots and histograms

If we download the ggExtra package we can combine our scatterplot with our previous exploratory boxplots.  Note that the syntax here is a big different.  We have to save the plot to an object (my.plot) and then use a function on that object: ggMarginal(my.plot).

Note: If this doesn't work, don't worry.

NOte: Don't ask for help getting this to work - its totally cherry on the top plotting.

```{r}
#load ggExtra
install.packages("ggExtra")
library(ggExtra)

#save the scatterplot

my.plot <- ggscatter(data = data.long,
          y = "height",
          x = "week",
          xlab = "Week of study",
          ylab = "Height of seedling",
          add = "reg.line",
          conf.int = TRUE) 

#make boxplot
ggMarginal(my.plot, type = "boxplot",margins = "y")

```

